\vspace{-0.5cm}The X-shooter GTO sample of GRB afterglow and host Galaxy spectra

### 2016-11-14 document/XSGRB_sample.tex
resolving conflict in pull #124

\usepackage{todonotes}

@@

	\cleardoublepage

	\section{Sample selection criteria}

@@

	A significant fraction of the bursts presented here \todo[inline]{Insert exact number} have already had their hosts investigated in \citet{Kruhler2015}, for which extractions of the hosts exist. The focus of the data presented here are on the afterglows themselves and in the absence of a clear afterglow, the host. We will not, however, investigate the hosts.

@@

	\cleardoublepage

	\section{Observations}

	\subsection{RRM observations} \label{RRM}
	The rapid-response mode is

	\section{Reduction scheme}

@@

	Additionally, the wavelength solution for all bursts have been re-calibrated. Using a synthetic sky spectrum \citep{Noll2012, Jones2013}, a first refinement of the wavelength solution have been obtained by cross-correlating with the observed sky.

	Both a multiplicative and an additive offset has been tested, but in terms of $\chi^2$, the model with only a multiplicative offset is preferred.

	\subsection{Spectral resolution}
	The afterglow spectra described in this paper are obtained in
	Target-of-Opportunity (override) mode.
	In most cases there is therefore little possibility to tweak slit widths to the
	seeing at the time of observations (i.e. to optimise spectral resolution and
	signal to noise), and almost all our data is therefore taken with a fixed set
	of slit widths and binning, described above.
	In a fair number of cases, the seeing full width at half maximum (FWHM) is
	considerably smaller than the silt width, and the delivered spectral resolution
	will then be determined by the seeing rather than slit width, as afterglows are
	point sources (this is evidently not the case for extended sources, e.g for
	host galaxies).
	The delivered resolution for slit width dominated spectra post-reduction and
	extraction can easily be determined from the bright sky emission lines.
	For afterglow spectra with very high signal to noise, the delivered spectral
	resolution can at times be determined from the science data themselves.
	However, in the presence of multiple velocity components in absorption, other
	forms of line broadening, and a lack of lines at some redshifts, this is
	difficult to do at poorer signal to noise ratios (the majority of spectra in
	our sample).
	A broad starting value for the expect resolution will help fitting of these
	spectra, and can be important in upper limit determination, and for this reason
	we construct a aim to construct a crude relation between the seeing and the
	delivered resolution at our slit width, binning, and reduction pipeline
	settings.
	To this end we use observations of telluric standard stars that are taken with
	identical instrument settings as our afterglow spectra, usually just after the
	science data, as part of the ESO X-shooter calibration plan.
	These spectra have been reduced together with the afterglow spectra, using
	identical pipeline settings with the same version of the pipeline.
	First we fit a Gaussian function in the spatial direction of the trace of the
	standard star at 792 nm (i.e. in the VIS arm).
	After this, we fit a series of  20 telluric absorption lines in the telluric
	standard star spectra with Gaussians, taking care to select transitions that
	are not almost-resolved multiples, should be intrinsically unresolved, and are
	in areas with well defined continuum flux.
	We pick 34 telluric standard stars spanning a range of DIMM seeing values, with
	the majority between $0.5-1.5 arcsec$.
	The resulting distribution of spectral FWHM (km/s) as a function of spatial
	FWHM at 792 nm is fairly well described by a linear relation $a + b*x$, with
	$x$ the spatial FWHM in pixels (with 0.15 arc sec per pixel),  $a= 21.4 \pm
	1.3$ km/s, $b=1.4 \pm 0.2$.
	We use this linear relation as a way to estimate the spectral resolution for
	medium to poor signal to noise afterglow spectra in the VIS arm.
	To extend this to the UVB and NIR arm, we measured a series of lines in NIR arm
	spectra of a subset of 19 sources used for the VIS arm above, and find that the
	resulting distribution is consistent with a simple scaling of the VIS arm
	relation by the ratio of resolutions of the NIR and VIS arm for unresolved,
	slit filling, sources as given on the ESO instrument website.
	The UVB arm contains no suitable absorption lines to use, and we therefore use
	a scaled value as in the NIR arm.
	While this simple method is not terribly accurate (for one, the spatial profile
	of the trace is not a perfect Gaussian), but it gives a sufficiently accurate
	estimate for the analysis of these poor signal to noise science spectra.

@@

	\subsection{Correction for offsets in the wavelength calibration}

	X-shooter, being installed at the VLT Cassegrain focusi is  prone to
	flexures during operations. The flexures modify the projection of the slit
	on the detector with respect to the one obtained in daytime calibration. 
	This require a modification of the wavelength solution in order to
	process correctly the night-time data. Part of this correction
	is performed by the pipeline using the frames taken
	during X-shooter Active Flexure Compensation procedure
	\footnote{X-shooter User Manual available at https://www.eso.org/sci/facilities/paranal/instruments/xshooter/doc.html}
	We corrected the remaining part using as a reference the sky
	emission lines present in the observed data.

@@

	\subsection{Properties of rejected triggers}
	How many bursts have been excluded due to particularities in the follow-up attempt. Visitor, Weather, technical issues.

@@

	\subsubsection{Comparison to X-ray determined columns}

@@

@@

	What is the expected transmission for a high-column HI, where the hydrogen primarily will be in the form of H2?

>>>>

\usepackage{xcolor}
\newcommand\todo[1]{\textcolor{red}{(#1)}}

@@

\subsection{Sample selection criteria and observations}

\subsection{Sample selection criteria} \label{samplecrit}

@@

A significant fraction of the bursts presented here \todo{Insert exact
number} have already had their hosts investigated in \citet{Kruhler2015}, for
which extractions of the hosts exist. The focus of the data presented here are
on the afterglows themselves and in the absence of a clear afterglow, the host.
We will not, however, investigate the hosts.

@@

\subsection{Follow-up procedure}

\subsubsection{RRM observations} \label{RRM}
The rapid-response mode is

\subsection{Observations}
All the observations in this sample has been carried out using the
cross-dispersed echelle spectrograph, X-shooter \citep{Vernet2011}, mounted on
two of the four Unit Telescopes at ESO/VLT, UT2 (Kueyen) and UT3 (Melipal)
during the duration of this follow-up campaign, which covers the entire lifetime
of X-shooter. Observations have been carried out from the ESO period P84
through P98 under the following programme IDs: 098.A-0055, 097.A-0036, 096.A-0079,
095.A-0045, 094.A-0134, 093.A-0069, 092.A-0124, 0091.C-0934, 090.A-0088,
089.A-0067, 088.A-0051, 087.A-0055, 086.A-0073, 085.A-0009 and 084.A-0260.

The first GRB followed up is GRB090313, observed 15th of March 2009, which was
while X-shooter was mounted on UT3 during the commissioning of the instrument.
The bursts observed during the commisioning period is not a part of our
statistical sample, but are nonetheless published as part of the bursts observed
by the X-shooter GRB team. The first burst observed after science verification
was completed when X-shooter was moved to UT2 is GRB091018 which thereby
constitute the first burst entering our statistical sample. For all bursts that
fulfill our sample selection criteria, described in Sect. \ref{samplecrit},
spectroscopic follow-up has been attempted with X-shooter. Various conditions
can affect our ability to follow up a given burst, and a discussion of this
effect is included in Sect. \ref{badbursts}.

X-shooter can cover the spectral wavelength region from 3000 \AA~to 24800 \AA~in
a single exposure, by splitting the light in three separate spectroscopic arms
using dichroics.
This way each arm work as a separate instrument, each functioning as its own
echelle spectrograph.
The ultraviolet blue(UVB) arm covering 3000 - 5500 \AA, the visual(VIS) arm
covering 5500 - 10200 \AA, and the near-infrared()NIR) arm covering 10200 -
24800 \AA~ with the possibility of applying a k-band
blocking filter cutting the coverage of the NIR arm at 10200 - 21000 \AA.

\todo{ABBA}

For the large majority of the bursts we have observed with a slit width of
1\farc0, 0\farc9, and 0\farc9 for the UVB, VIS, and NIR-arm respectively, which
puts a lower limit on the delivered resolution of the spectra based on the
tabulated values of the delivered resolutions, which is 4350, 7450, and 5300 for
the UVB, VIS and NIR-arm respectively\footnote{\url{https://www.eso.org/sci/facilities/paranal/instruments/xshooter/inst.html}}.
The slit width also sets the width of the atmospheric sky lines and determines
the amount of light lost due to the wavelength-dependent seeing PSF extending
outside the coverage of the slit, where the width of sky-lines is always set by
the slit width whereas both the delivered resolution and the slitloss changes
for the better as the seeing PSF drops below the slit width.
For atmospheric conditions delivering a seeing PSF with a FWHM of 0\farc9
observed with a 0\farc9 slit only 76.1 percent of the light will enter the slit,
meaning that for almost all observations a slitloss correction is required. We
describe how slitlosses were corrected for in Sect. \ref{postproc}.
For accurate measurements of velocity widths, a precise instrumental resolution
is required and this becomes better when the delivered seeing is better than the
slit width. We discuss this in Sect. \ref{resolution}.

\todo{ADC discussion}

\section{Data processing}

In this section we describe how the final data products are produced.

Before any reductions are done, the raw object images are run through the
cosmic-ray removal algorithm \citep{VanDokkum2001} implementation,
\textit{Astro-SCRAPPY} \footnote{\url{https://github.com/astropy/astroscrappy}},
where a wide clipping radius have been used around detected cosmics to ensure that edge residuals are
robustly rejected.

The basis for the reductions is the VLT/X-shooter pipeline, version
\texttt{2.7.1} or newer \citep{Goldoni2006, Modigliani2010}. The pipeline is
managed with the Reflex interface \citep{Freudling2013} and is used for
subtraction of bias level, flat-fielding, tracing of the echelle orders,
wavelength calibrations with the use of arc-line lamps, flux calibration using
spectrophotometric standards \citep{Vernet2010, Hamuy1994}, mirror flexure
compensation, initial sky-subtraction and lastly the rectification and merging
of the orders. For the initial sky-subtraction, the background has been
estimated in regions adjacent to the object trace clear of contaminating
sources. Because of the broken ADC, for some objects there is a lot of curvature
in the object trace. This means that for some bursts, the initial sky-estimate
has been made from a limited number of pixels in the spatial direction. By doing
an initial subtracting the sky on the un-rectified image we ensure that bulk of
the sky background is not redistributed by the rectification process.

The image is rectified onto an equidistant grid with a dispersion sampling of
0.2 \AA/pixel and a 0.16 "/pixel spatial sampling for the UVB and VIS arm and
0.6 \AA/pixel with a 0.21 "/pixel in the NIR arm.  Because the tabulated
resolution is a lower limit, by choosing a sampling of 0.2 \AA/pixel, we ensure
that the bluest part of neither of the arms have a sampling lower than the
Nyquist sampling rate of 2 pixels per resolution FWHM.

\subsection{Post-processing} \label{postproc}

For a typical observation, each of the exposures in the nodding sequence has
been reduced as single observation and then subsequently combined to form a
single image. Because this strategy is employed, we can reject outliers in the
stack and weight by the average inverse variance of the background. When
weighting images where the noise in each pixel is dominated by Poisson noise it
important to estimate the background variance in a large enough region, so that
to remove the correlation between the signal and the weights. To this end, the
weight map is generated by a running median window over the variance map, where
the trace as been masked and width of the window is chosen to be wide enough for
median variance to be generated on the basis of several hundred pixels. This
weighting scheme automatically also optimally combines images of different
exposure times or images where the background is varying, which is often the
case when a burst has been observed close to twilight

Because the background is very bright and there is a high abundance of broad
sky-lines in the NIR arm, when there are no contaminating sources in the slit,
the sky has been put back on the images and they have been reduced in pairs of
two, subtracting the two from each other, keeping the WCS static. This amounts
to the regular nodding reduction, only we can reject outliers and weight by the
averaged inverse variance map.

\subsection{Correction for offsets in the wavelength calibration}    \label{wavecal}

Additionally, the wavelength solution for all bursts have been re-calibrated. Using a synthetic sky spectrum \citep{Noll2012, Jones2013}, a first refinement of the wavelength solution have been obtained by cross-correlating with the observed sky. 
Both a multiplicative and an additive offset has been tested, but in terms of $\chi^2$, the model with only a multiplicative offset is preferred.

X-shooter, being installed at the VLT Cassegrain focus is  prone to
flexures during operations. The flexures modify the projection of the slit
on the detector with respect to the one obtained in daytime calibration. 
This require a modification of the wavelength solution in order to
process correctly the night-time data. Part of this correction
is performed by the pipeline using the frames taken
during X-shooter Active Flexure Compensation procedure
\footnote{X-shooter User Manual available at https://www.eso.org/sci/facilities/paranal/instruments/xshooter/doc.html}
We corrected the remaining part using as a reference the sky
emission lines present in the observed data.

\subsection{Spectral extraction}    \label{extract}

@@

\subsection{Spectral resolution} \label{resolution}
The afterglow spectra described in this paper are obtained in
Target-of-Opportunity (override) mode.
In most cases there is therefore little possibility to tweak slit widths to the
seeing at the time of observations (i.e. to optimise spectral resolution and
signal to noise), and almost all our data is therefore taken with a fixed set
of slit widths and binning, described above.
In a fair number of cases, the seeing full width at half maximum (FWHM) is
considerably smaller than the silt width, and the delivered spectral resolution
will then be determined by the seeing rather than slit width, as afterglows are
point sources (this is evidently not the case for extended sources, e.g for
host galaxies).
The delivered resolution for slit width dominated spectra post-reduction and
extraction can easily be determined from the bright sky emission lines.
For afterglow spectra with very high signal to noise, the delivered spectral
resolution can at times be determined from the science data themselves.
However, in the presence of multiple velocity components in absorption, other
forms of line broadening, and a lack of lines at some redshifts, this is
difficult to do at poorer signal to noise ratios (the majority of spectra in
our sample).
A broad starting value for the expect resolution will help fitting of these
spectra, and can be important in upper limit determination, and for this reason
we construct a aim to construct a crude relation between the seeing and the
delivered resolution at our slit width, binning, and reduction pipeline
settings.
To this end we use observations of telluric standard stars that are taken with
identical instrument settings as our afterglow spectra, usually just after the
science data, as part of the ESO X-shooter calibration plan.
These spectra have been reduced together with the afterglow spectra, using
identical pipeline settings with the same version of the pipeline.
First we fit a Gaussian function in the spatial direction of the trace of the
standard star at 792 nm (i.e. in the VIS arm).
After this, we fit a series of  20 telluric absorption lines in the telluric
standard star spectra with Gaussians, taking care to select transitions that
are not almost-resolved multiples, should be intrinsically unresolved, and are
in areas with well defined continuum flux.
We pick 34 telluric standard stars spanning a range of DIMM seeing values, with
the majority between $0.5-1.5 arcsec$.
The resulting distribution of spectral FWHM (km/s) as a function of spatial
FWHM at 792 nm is fairly well described by a linear relation $a + b*x$, with
$x$ the spatial FWHM in pixels (with 0.15 arc sec per pixel),  $a= 21.4 \pm
1.3$ km/s, $b=1.4 \pm 0.2$.
We use this linear relation as a way to estimate the spectral resolution for
medium to poor signal to noise afterglow spectra in the VIS arm.
To extend this to the UVB and NIR arm, we measured a series of lines in NIR arm
spectra of a subset of 19 sources used for the VIS arm above, and find that the
resulting distribution is consistent with a simple scaling of the VIS arm
relation by the ratio of resolutions of the NIR and VIS arm for unresolved,
slit filling, sources as given on the ESO instrument website.
The UVB arm contains no suitable absorption lines to use, and we therefore use
a scaled value as in the NIR arm.
While this simple method is not terribly accurate (for one, the spatial profile
of the trace is not a perfect Gaussian), but it gives a sufficiently accurate
estimate for the analysis of these poor signal to noise science spectra.

@@

@@

\subsection{Properties of rejected triggers} \label{badbursts}
How many bursts have been excluded due to particularities in the follow-up attempt. Visitor, Weather, technical issues.

@@

@@

\begin{figure}
	\centerline{\includegraphics[width=10cm]{figures/OAXR_NH_sphere.pdf}}
	\caption{Optically derived HI column densities vs. X-ray derived ones taken from \citet{Buchner2016}. The colors of the points reflect the redshift of the burst. Circles mark the column densities published in \citet{Fynbo2009} and squares mark the ones derived as part of the sample presented here. Overplot in blue is the 1:1 correspondence, expected if the conditions local to the GRBs are similar to the local ISM abundances. If the metallicity is lower at the GRB site, the point move upwards in the plot. }
	\label{fig:OAXR}
\end{figure}

@@

What is the expected transmission for a high-column HI, where the hydrogen primarily will be in the form of H2? Number of bursts with detected H2 and their HI column density. Rahmati Schaye, Krongold Prochaska.
